---
title: "Distribution at the Nodes"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r include=FALSE}
knitr::opts_chunk$set(tidy=FALSE,fig.width=5.5,fig.height=3.5, cache = TRUE)
library(tigerstats)
library(tigerTree)
```

## Introduction

Sometime we build a classification tree, not in order to make a prediction for an individual, but rather to identify "groups" of individuals that have an elevated probability of possessing a particular characteristic.

For example, suppose that in the case of Justin Verlander's pitches we would like to say which sorts of pitches are especially likely to be curve balls (value "CU").

First let's set up the data frame:

```{r}
ver2 <- verlander
ver2$season <- NULL
ver2$gamedate <- NULL
```

Now we divide the data into training and test sets:

```{r}
dfs <- divideTrainTest(seed = 3030, prop.train = 0.5, data = ver2)
verTrain <- dfs$train
verTest <- dfs$test
```

Then we build a tree on the training set:

```{r}
trMod <- tree(pitch_type ~ ., data = verTrain)
trMod
```

We can tell that pitches at terminal nodes 10 and 23 are especially likely to be curve balls.

## Distribution at the Nodes:  Taining Set

The `distAtNodes()` function gives us this information directly, in the form of a table:

```{r}
trainTab <- distAtNodes(trMod, df = verTrain, resp_varname = "pitch_type")
trainTab
rowPerc(trainTab)
```

## Distribution at the Nodes:  Taining Set

The model points to two nodes as having "high-likelihood nodes" for curve-balls.  But a more reliable estimate of the probability of a curve-ball at each of these nodes  is provided by applying the model to *new* data, not the data that is used to build the model itself.  Hence we also use the `distAtNodes()` function on the model and the *test* set:

```{r}
testTab <- distAtNodes(trMod, df = verTest, resp_varname = "pitch_type")
testTab
rowPerc(testTab)
```

Again nodes 10 and 23 stand out---and with about the same probability of curve-ball at each node.

## Caution

When a node does not contain a large number of individuals, estimates of distribution of the response variable at that node are subject to a great deal of chance variation.  Watch for this especially when you are building trees with many nodes, or when your training or test sets are not large to being with. 










